An introduction to Docker for reproducible research, with examples from the R environment

### 2014-09-16 manuscript.Rmd
writing

Reproducible research has recieved an increasing level of attention
throughout the scientific community (e.g. [Peng 2011], [Nature editorial])
and the public at large (e.g. [The Economist 2014]). All steps of the
scientific process, from data collection and processing, to analyses,
visualizations and conclusions depend ever more on computation and
algorithms, _computational reproducibility_ has recieved particular
attention [Merali 2010].  Though in principle this algorithmic dependence
should make such research easier to reproduce -- computer codes being
both more portable and potentially more precise to exchange and run
than experimental methods -- in practice this has led to an ever larger
and more complex black box that stands between what was actually done
and what is described in the literature.  Crucial scientific processes
such as replicating the results, extending the approach or testing the
conclusions in other contexts, or even merely installing the software
used by the original researchers can become immensely time-consuming if
not impossible.

@@

<!-- FIXME more stuff on what this means-->

@@

<!-- Figure 1: technical reproducibility stack
1. Operating system
2. Workflow systems 
3. Package managers 
4. Dynamic documents

@@

These issues are reflected both during the original research or development
phase and in any subsequent reuse. Another researcher may be less likely
to build on existing work if it can only be done by using a particular
workflow system or monolothic software platform with which they are
unfamiliar.  Likewise, a user is more likey to make their own computational
environment avialable for reuse if it does not involve a significant
added effort in packaging and documenting [Stodden 2010].

@@

<!---
- The ability to work in a familiar, local environment (during one's own research or when re-using elements from another researcher)
- An Docker container approach to both development and deployment
- Image re-use & Image linking 
-->

@@

Perhaps the most important feature of a reproducible research tool is
that it be easy to learn and fit relatively seamlessly into exsiting
workflow patterns of domain researchers.  As we have seen, this, more than
any other concern, can explain the relatively low uptake of previously
proposed solutions.  <!-- too strong? -->

Being a new and unfamiliar tool to most domain scientists, Docker is
far from immune to the same critique.  Nevertheless, Docker takes us
several key steps towards an approach that can be easily adopted in a
research context.

Primary among these is the ability to run Docker locally on all major
operating systems with minimal prior expertise.  On systems not already
based on the Linux Kernel (such as Mac or Windows clients), this is
accomplished through the use of a small VirtualBox-based VM running on
the host OS called `boot2docker`.

Local development offers several important advantages in encouraging
adoption.

- Shared volumes
- Host OS tools

-----

In the R environment, the open source RStudio IDE provides
another key component in making this system accessible to most
domain scientists. Because the RStudio IDE is written completely with
standard web languages (javascript, css, and html), its web-based 
RStudio Server looks and feels identical to its popular desktop IDE.
RStudio Server provides a way to interact with R on a remote environment
without the latency, X tunnelling, etc typically involved in running R
on a remote server.

For our purposes, it provides users a way to interact with R,
the filesystem, git, text editors, and graphics running on a Docker
container.  Users already familiar with the popular IDE can thus benefit
from the reproduciblity and portability features provided by running R
in a container environment without having to adapt to a new workflow
for managing these things that is usually necessary when moving to a
cloud environment.

RStudio-Server is not the only way to provide a familiar developer
environment to users working with a development environment running in
a container on their local computer.  Docker supports the sharing of
volumes (directories) between the container and the host OS.  This allows
a user to rely on the familiar tools of the host OS for roles such as
text editing, file browsing, or version control, while still allowing
code execution to occur inside the controlled development environment
of the container. In this example, we launch an interactive R console
in a container that is linked to our current working directory:

@@

(Clearly a similar command could be used just as well with interactive
shells from other languages such as `ipython` or `irb`). The resulting
system behaves almost identically[^2] to running R on the command line
of the host OS.

@@

@@

First, Docker makes it immensely easy to build one container on top of
another.  Rather than copy a Dockerfile and then start adding more lines
to the bottom, we can declare a new Dockerfile is built on an old one
using the `FROM` directive. Here is an example Dockerfile that adds the
R statistical computing environment to a basic Ubuntu Linux distribution.

@@

<!-- too much detail? -->
The implementation of this would nonetheless depend on the individual 
researcher's willingness to break things into multiple containers, as
it is indeed possible to run everything in a monolithic container. 
This is not necessarily bad, as it simplifies the workflow (no need
to execute and link separate containers), and can be refactored into
multiple containers when appropriate. It remains to be seen whether
linking containers through this mechanism

@@

- Interactive use, scripted use (Dockerfiles)
- Reverting to a previous state
- Integration with local development environment
- Integration with cloud computing
- Transparent & extensible
- serial and linking containers
- Best practices are still taking shape

----------

@@

3) You can run the docker container *locally*.  I think this is huge.
In my experience, most researchers do their primary development
locally.  By running RStudio-server on your laptop, it isn't necessary
for me to spin up an EC2 instance (with all the knowledge & potential
cost that requires).  By sharing directories between Docker and the
host OS, a user can still use everything they already know -- their
favorite editor, moving files around with the native OS
finder/browser, using all local configurations, etc, while still
having the code execution occur in the container where the software is
precisely specified and portable.  Whenever you need more power, you
can then deploy the image on Amazon, DigitalOcean, a bigger desktop,
your university HPC cluster, your favorite CI platform, or wherever
else you want your code to run.  (On Mac & Windows, this uses
something called boot2docker, and was not very seamless early on.  It
has gotten much better and continues to improve.)

4) Versioned images.  In addition to version managing the Dockerfile,
the images themselves are versioned using a git-like hash system
(check out: docker commit, docker push/pull, docker history, docker
diff, etc).  They have metadata specifying the date, author, parent
image, etc.  We can roll back an image through the layers of history
of its construction, then build off an earlier layer.  This also
allows docker to do all sorts of clever things, like avoiding
downloading redundant software layers from the docker hub.  (If you
pull a bunch of images that all build on ubuntu, you don't get n
copies of ubuntu you have to download and store).  Oh yeah, and
hosting your images on Docker hub is free (no need to pay for an S3
bucket... for now?) and supports automated builds based on your
dockerfiles, which acts as a kind of CI for your environment.
Versioning and diff'ing images is a rather nice reproducibility
feature.

--------------------------------

<!-- Covered elsewhere in a more integrated fashion -->
Researchers familiar with LXC may be tempted to view the Docker
system as little more than a creative application and marketing of an
existing technology. However, the Docker approach adds several features
crucial for its use in reproducible research (adapted from [Docker
FAQ](https://docs.docker.com/faq/#what-does-docker-add-to-just-plain-lxc)):

@@

- NSF
- rOpenSci
- Dirk Eddlebuettel, Rich FitzJohn, Yihui Xie, Titus Brown, John Stanton-Geddes and others for helpful discussions about reproducibility, virtualization, and Docker that have helped shape this manuscript.

@@

[Garijo 2013]:     http://doi.org/10.1371/journal.pone.0080278
[Lapp 2014]: https://storify.com/hlapp/reproducibility-repeatability-bigthink
[Gilbert 2012]: 10.1111/j.1365-294X.2012.05754.x
[Collberg 2014]: http://reproducibility.cs.arizona.edu/v1/tr.pdf
[Brown 2014]: http://cs.brown.edu/~sk/Memos/Examining-Reproducibility/

[Vandewalle 2009]: http://doi.org/10.1109/MSP.2009.932122
[Merali 2010]: http://doi.org/10.1038/467775a "Computational Science ... error: why scientific programming does not compute"
[Gil 2007]: http://doi.org/10.1109/MC.2007.421 "Examining the Challenges of Scientific Workflows"
[Hull 2006]: http://doi.org/10.1093/nar/gkl320 "Taverna: a tool for building and running workflows of services"
[Altintas 2004]: http://doi.org/10.1109/SSDM.2004.1311241 "Kepler: an extensible system for design and execution of scientific workflows"
[Howe 2012]: http://doi.org/10.1109/MCSE.2012.62
[Dudley 2013]: http://doi.org/10.1038/nbt1110-1181 
[Clark 2014]: https://berkeley.app.box.com/s/w424gdjot3tgksidyyfl
[Ooms 2013]: http://arxiv.org/abs/1303.2140v2
[Ooms 2014]: http://arxiv.org/abs/1406.4806
[Peng 2011]: http://doi.org/10.1126/science.1213847
[Barnes 2010]: http://doi.org/10.1038/467753a
[Ince 2012]: http://doi.org/10.1038/nature10836
[Stodden 2014]: http://doi.org/10.3233/SJI-140818
[Stodden 2010]: http://doi.org/10.2139/ssrn.1550193
[Stodden 2013]: http://www.davidhbailey.com/dhbpapers/icerm-report.pdf "ICERM Working group report: Setting the Default to Reproducible: Reproducibility in Computational and Experimental Mathematics"

[Nature editorial]: http://doi.org/10.1038/483509a
[Nature Special Issue]: http://www.nature.com/nature/focus/reproducibility/

[van der Meulen 2008]: http://doi.org/10.1109/TSE.2008.70 "The effectiveness of software diversity in a large population of programs."

[Watson 2012]: https://twitter.com/BioMickWatson/status/265037994526928896

>>>>

Reproducible research has recieved an increasing level of attention
throughout the scientific community (e.g. [Peng 2011], [Nature 2014])
and the public at large (e.g. [The Economist 2013]). All steps of the
scientific process, from data collection and processing, to analyses,
visualizations and conclusions depend ever more on computation and
algorithms, _computational reproducibility_ has recieved particular
attention [Merali 2010].  Though in principle this algorithmic dependence
should make such research easier to reproduce -- computer codes being
both more portable and potentially more precise to exchange and run
than experimental methods -- in practice this has led to an ever larger
and more complex black box that stands between what was actually done
and what is described in the literature.  Crucial scientific processes
such as replicating the results, extending the approach or testing the
conclusions in other contexts, or even merely installing the software
used by the original researchers can become immensely time-consuming if
not impossible.

@@

@@

<!-- Figure 1: technical reproducibility stack
1. Operating system
2. Workflow systems 
3. Package managers 
4. Dynamic documents
-->

@@

These issues are reflected both during the original research or development
phase and in any subsequent reuse. Another researcher may be less likely
to build on existing work if it can only be done by using a particular
workflow system or monolothic software platform with which they are
unfamiliar.  Likewise, a user is more likey to make their own computational
environment avialable for reuse if it does not involve a significant
added effort in packaging and documenting ([Stodden 2010]).

@@

@@

Perhaps the most important feature of a reproducible research tool is
that it be easy to learn and fit relatively seamlessly into exsiting
workflow patterns of domain researchers.  As we have seen, this,
more than any other concern, can explain the relatively low uptake
of previously proposed solutions.  Being a new and unfamiliar tool to
most domain scientists, Docker is far from immune to the same critique.
Nevertheless, Docker takes us several key steps towards an approach that
can be easily adopted in a research context.

While proponents of virtual machines for reproducible research propose
that these machines would be available exclusively as cloud computing
environments (e.g. [Dudley 2014]), many researchers work _locally_, that
is, primarily with software that is installed on their laptop or desktop
computer, and turning to cloud-based or other remote platforms only for
certain collaborative tasks or when the work is mature enough to need
increased computational power.  Working locally allows a researcher to
rely more on the graphic interface tools for tasks such as managing
files, text editing, debugging, IDEs, or interacting with version
control systems.  Scientific computing on remote machines, by contrast,
still relies largely on potentially less familiar text based command
line functions for these tasks (though web based interfaces are rapidly
filling this gap).

Docker can be easily installed on most major platforms (see
[https://docs.docker.com/installation](https://docs.docker.com/installation);
On systems not already based on the Linux Kernel, such as Mac or Windows,
this is accomplished through the use of a small VirtualBox-based VM
running on the host OS called `boot2docker`) and run locally.  Docker
allows a user to link any directory (such as the working directory
for a particular script or project) to the running Docker container.
This allows a user to rely on the familiar tools of the host OS for
roles such as text editing, file browsing, or version control, while
still allowing code execution to occur inside the controlled development
environment of the container.

For example, we launch an interactive R console in a container that
is linked to our current working directory:

@@

The resulting system behaves almost identically[^2] to running R on
the command line of the host OS. (Clearly a similar command could be
used just as well with interactive shells from other languages such as
`ipython` or `irb`).

@@

### Using RStudio Server

An alternative approach for working locally with familar tools is to
leverage web-based clients such as RStudio. 
In the R environment, the open source RStudio IDE provides
another key component in making this system accessible to most
domain scientists. Because the RStudio IDE is written completely with
standard web languages (javascript, css, and html), its web-based 
RStudio Server looks and feels identical to its popular desktop IDE.
<!--
RStudio Server provides a way to interact with R on a remote environment
without the latency, X tunnelling, etc typically involved in running R
on a remote server. -->

RStudio server provides users a way to interact with R,
the filesystem, git, text editors, and graphics running on a Docker
container.  Users already familiar with the popular IDE can thus benefit
from the reproduciblity and portability features provided by running R
in a container environment without having to adapt to a new workflow.

To accompany this paper, I provide a Docker image for running RStudio
server, which can be launched as follows.  From the terminal (or
boot2docker terminal on Mac or Windows client), run:

```bash
sudo docker run -d -p 8787:8787 cboettig/ropensci
```

That will take a while to download the image the first time you run
it. boot2docker users will then need to determine the ip address assigned
to boot2docker, while Linux users can just specify `http://localhost`

```bash
boot2docker ip
```

Add the port `:8787` to the end of this address and paste
it into your browser address bar, which should open to the
RStudio welcome screen.  A user can now login with user/password
`rstudio/rstudio`, run R scripts, install packages, use git,
and so forth. User login and other configurations can be
customized using environmental variables; see details at
[https://github.com/ropensci/docker](https://github.com/ropensci/docker)

### Portable computation

A particular advantage of this approach is that the resulting
computational environment is immediately _portable_.  This is useful not
only for the purposes of reproducible research, where other users may seek
to reconstruct the computational environment necessary to run the code,
but is also of immediate value to the researcher themselves. For instance,
a researcher might want to execute their code on a cloud server which
has more memory or processing power then their local machine, or would
want a co-author to help debug a particular problem.  In either case,
the researcher can export a snapshot of their running container:

```bash
docker export <container-name> > container.tar
```
and then run this identical environment on the cloud or collaborators' machine.

@@

First, Docker facilitates modular reuse by build one container on top of
another.  Rather than copy a Dockerfile and then start adding more lines
to the bottom, we can declare a new Dockerfile is built on an old one
using the `FROM` directive. Here is an example Dockerfile that adds the
R statistical computing environment to a basic Ubuntu Linux distribution.

@@

<!-- Transition -->

## Versioning

<!-- FIXME: Email text.  revise -->

 In addition to version managing the Dockerfile,
the images themselves are versioned using a git-like hash system
(e.g. see `docker commit`, `docker push`/`docker pull`, `docker history`, 
`docker diff`).  Docker images and containers have dedicated metadata specifying the date, author, parent
image, and other details (see `docker inspect`).  We can roll back an image through the layers of history
of its construction, then build off an earlier layer, or roll back changes
we have made interactively in a container. For instance, here we can
inspect recent changes made to the `ubuntu:latest` image:

```bash
docker history ubuntu:latest
```

We can identify an earlier version, and roll back to that version 
just by adusting the docker tag to match the hash of that version. 
For instance:

```bash
docker tag 25f ubuntu:latest
```
If we now inspect the history, we see it begins from this earlier poitn:

```bash
docker history ubuntu:latest
```

This same feature also means that docker can perform incremential
uploads and downloads that send only the differences between images,
(just like `git push` or `git pull` for git repositories), rather than
transfer the full image each time.

<!-- Step back, big picture -->

@@

<!--
The implementation of this would nonetheless depend on the individual 
researcher's willingness to break things into multiple containers, as
it is indeed possible to run everything in a monolithic container. 
This is not necessarily bad, as it simplifies the workflow (no need
to execute and link separate containers), and can be refactored into
multiple containers when appropriate. 
-->

@@

@@

CB acknowledges support from NSF grant DBI-1306697, and also from the Sloan Foundation support through the rOpenSci project. CB also wishes to thank Dirk Eddlebuettel, Rich FitzJohn, Yihui Xie, Titus Brown, John Stanton-Geddes and many others for helpful discussions about reproducibility, virtualization, and Docker that have helped shape this manuscript.

@@

[Garijo 2013]: http://doi.org/10.1371/journal.pone.0080278
[Lapp 2014]: https://storify.com/hlapp/reproducibility-repeatability-bigthink
[Gilbert 2012]: 10.1111/j.1365-294X.2012.05754.x
[Collberg 2014]: http://reproducibility.cs.arizona.edu/v1/tr.pdf
[Brown 2014]: http://cs.brown.edu/~sk/Memos/Examining-Reproducibility/
[Vandewalle 2009]: http://doi.org/10.1109/MSP.2009.932122
[Merali 2010]: http://doi.org/10.1038/467775a "Computational Science ... error: why scientific programming does not compute"
[Gil 2007]: http://doi.org/10.1109/MC.2007.421 "Examining the Challenges of Scientific Workflows"
[Hull 2006]: http://doi.org/10.1093/nar/gkl320 "Taverna: a tool for building and running workflows of services"
[Altintas 2004]: http://doi.org/10.1109/SSDM.2004.1311241 "Kepler: an extensible system for design and execution of scientific workflows"
[Howe 2012]: http://doi.org/10.1109/MCSE.2012.62
[Dudley 2013]: http://doi.org/10.1038/nbt1110-1181 
[Clark 2014]: https://berkeley.app.box.com/s/w424gdjot3tgksidyyfl
[Ooms 2013]: http://arxiv.org/abs/1303.2140v2 "Possible Directions for Improving Dependency Versioning in R"
[Ooms 2014]: http://arxiv.org/abs/1406.4806 "The OpenCPU System: Towards a Universal Interface for Scientific Computing through Separation of Concerns"
[Peng 2011]: http://doi.org/10.1126/science.1213847
[Barnes 2010]: http://doi.org/10.1038/467753a
[Ince 2012]: http://doi.org/10.1038/nature10836
[Stodden 2014]: http://doi.org/10.3233/SJI-140818
[Stodden 2010]: http://doi.org/10.2139/ssrn.1550193
[Stodden 2013]: http://www.davidhbailey.com/dhbpapers/icerm-report.pdf "ICERM Working group report: Setting the Default to Reproducible: Reproducibility in Computational and Experimental Mathematics"
[Nature 2014]: http://doi.org/10.1038/483509a
[The Economist 2013]: http://www.economist.com/news/leaders/21588069-scientific-research-has-changed-world-now-it-needs-change-itself-how-science-goes-wrong
[van der Meulen 2008]: http://doi.org/10.1109/TSE.2008.70 "The effectiveness of software diversity in a large population of programs."
[Watson 2012]: https://twitter.com/BioMickWatson/status/265037994526928896

```{r include=FALSE}
library("knitcitations")
options(citation_format = "pandoc")

citep(c(
"10.1371/journal.pone.0080278",
"10.1126/science.1213847",
"10.1038/467753a",
"10.1038/nature10836",
"10.1038/483509a",
"10.1038/467775a",
"10.1111/j.1365-294X.2012.05754.x",
"10.1109/MSP.2009.932122",
"10.1109/MC.2007.421",
"10.1093/nar/gkl320"))

citep(c(
"10.1109/SSDM.2004.1311241",
"10.1109/MCSE.2012.62"))

citep("10.1038/nbt1110-1181")

citep("http://doi.org/10.3233/SJI-140818")

citep(c(
"10.2139/ssrn.1550193",
"10.1109/TSE.2008.70"))

citep(c(
"http://arxiv.org/abs/1303.2140v2",
"http://arxiv.org/abs/1406.4806",
"https://storify.com/hlapp/reproducibility-repeatability-bigthink",
"http://cs.brown.edu/~sk/Memos/Examining-Reproducibility/",
"https://berkeley.app.box.com/s/w424gdjot3tgksidyyfl",
"http://reproducibility.cs.arizona.edu/v1/tr.pdf",
"http://www.davidhbailey.com/dhbpapers/icerm-report.pdf",
"http://www.economist.com/news/leaders/21588069-scientific-research-has-changed-world-now-it-needs-change-itself-how-science-goes-wrong",
"https://twitter.com/BioMickWatson/status/265037994526928896"))

write.bibtex(file="components/references.bib")

```

<!-- Covered elsewhere in a more integrated fashion -->
<!--

Researchers familiar with LXC may be tempted to view the Docker
system as little more than a creative application and marketing of an
existing technology. However, the Docker approach adds several features
crucial for its use in reproducible research (adapted from [Docker
FAQ](https://docs.docker.com/faq/#what-does-docker-add-to-just-plain-lxc)):
